---
title: "Subsetting Data"
author: "JLB"
date: "2023-08-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Subsetting Data in R

- Subsequent data processing setps (statistical analysis, graph generation) works best if you select the data that you are interested in. 

- Splitting up large data sets into multiple subsetted dataframes has many functions in R programming and is a useful tactic to maintain readable code blocks. 

# Get data

### [Link to explaination about raw data](https://black-cat-enthusiast.github.io/Purple_Monkey_Dishwasher/2023/07/12/example-data/)

### [Link to raw data](https://github.com/black-cat-enthusiast/Blogdown_Test/blob/master/content/post/2023-07-12-Example-Data/EB_Rats_Nicotine_Sensitization.csv)

```{r}
library(tidyverse) # Load the tidyverse package
data <- read_csv("EB_Rats_Nicotine_Sensitization.csv") # Call in the data
head(data) # check out the first six rows of data. 
```

# Subsetting = Select only a portion of the data. 

There are multiple methods to accomplish selecting only some of the data in R. 3 different approaches are outlined below. Which approach is preferable for a given situation depends on the parameters of the task and the desired output. 

### Method #1: by position 

- In R language, square brackets mean "where". Inside of the square brackets, there can be two sets of numbers separated by a comma. 
    + Values that come before the comma refer to the ROWS 
    + Values that come after the comma refer to the COLUMNS 
    + A blank space means "take all" 

```{r}
# Select all rows of the first 6 columns 
data[ ,1:6]
```

```{r}
# Select all columns of the first 6 rows 
data[1:6, ]
```

- You can also assign the subsetted data frame onto a new variable 
- It is generally better to create new variables than to overwrite existing variables. 

```{r}
data_2 <- data[ ,1:6] # Assign the first 6 columns of "data" onto a new variable named data_2
head(data_2) # Print out the first 6 rows of data_2
```

```{r}
# Select the first 6 rows of the first 6 columns 
data[1:6,1:6]
```

### Method #2: by content

- Especially when working with larger datasets, you may want to subset based on cell values rather than position within the dataframe. Selecting based on content is a more robust approach because it does not require manual accuracy checks. 

- Including a comma and a space after the expression instructs R to take all the data from the columns. This is important as the program will throw an error if you forget to include comma space.  

```{r}
# Select all columns for the rows that have the value "OIL" in the column PREhorm:
data[data$PREhorm == 1, ] # Select instances of "data" where the value in the column PREhorm equals exactly 1, take all rows
```

Data can also be selected based on multiple criteria:

```{r}
# Select all columns for rows where PREhorm equals exactly 1 AND CHALhorm equals exactly 1. 
data[data$PREhorm == 1 & data$CHALhorm == 1, ]
```

### Method #3: dplyr "select"

- You can also use the "select" function inside of dplyr code blocks. This is generally the most elegant solution to choosing a subset of your dataframe to work with.

- The result of the "select" function inside dplyr code blocks does not need to be assigned onto another variable, and the result of the below expression could be piped directly to the summarise function and / or to ggplot2 for visualization. 

```{r}
data %>%
  select(c("ID","PREhorm","Hab"))
```





