---
title: "ANOVA "
author: "JLB"
date: "2023-07-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE,
                      warning = FALSE)

library(tidyverse)
library(ggpubr)
library(reshape2)
options(scipen = 999)
```

# The General Linear Model

The GLM is a *family* of analytic approaches that involves fitting lines to data points.

### The underlying formula for the GLM: 


$$ \dfrac{signal}{noise} $$

OR, in other words,

$$ \dfrac{systematic\ variance}{error\ variance} $$


# ANOVA (Analysis of Variance)

Like [t-tests](), ANOVA **requires** *categorical IV(s)* and a *continuous DV*

- ANOVA can handle any number of groups and any number of factors. 
  + This is the main advantage of ANOVA over t-tests, which can only handle **2 groups** and **a single factor**

# Dimensions of the ANOVA: 

### The number of factors: 

- **"One-Way ANOVA"** = 1 factor
- **"Two-way ANOVA"** = 2 factors

### The experimental design:

- **repeated measures** = Multiple measurements involving the same individuals.
- **independent samples** = Comparisons involving multiple groups of individuals.

### [Link to explaination about data used in the examples below](https://black-cat-enthusiast.github.io/Purple_Monkey_Dishwasher/2023/07/12/example-data/)

### [Link to raw data](https://github.com/black-cat-enthusiast/Purple_Monkey_Dishwasher/blob/master/content/post/2023-08-12-Example-Data/EB_Rats_Nicotine_Sensitization.csv)

```{r, include=FALSE}
library(tidyverse)
data <- read_csv("EB_Rats_Nicotine_Sensitization.csv")


data$PREhorm = factor(data$PREhorm,# change the PREhorm variable to a factor
                      levels = c(0,1), # With two levels, 0 and 1 (ordered)
                      labels = c("OIL","EB")) # And label those levels, the condition names.

data$CHALhorm = factor(data$CHALhorm, # change the PREhorm variable to a factor
                      levels = c(0,1),
                      labels = c("OIL","EB"))

data <- data %>%
  mutate(group = case_when(
    data$PREhorm == "OIL" & data$CHALhorm == "OIL" ~ "OIL / OIL",
    data$PREhorm == "EB" &  data$CHALhorm == "OIL" ~ "EB / OIL",
    data$PREhorm == "OIL" & data$CHALhorm == "EB" ~ "OIL / EB",
    data$PREhorm == "EB" &  data$CHALhorm == "EB" ~ "EB / EB"
  ))

data$group <- as.character(data$group)
data$group <- factor(data$group, levels=c("OIL / OIL", "EB / OIL", "OIL / EB", "EB / EB"))

my_sum <- data %>%
  group_by(group) %>%
  summarise(n=n(),
            mean=mean(CHAL),
            sd=sd(CHAL)) %>%
  mutate(se=sd / sqrt(n))

my_colours <- c("blue","red","green","purple")

x <- ggplot(my_sum, aes(x=group,y=mean,colour=group,fill=group))+
  geom_bar(stat="identity", alpha=0.2)+
  geom_errorbar(aes(x=group,ymin=mean-se,ymax=mean+se, width=0.5))+
  scale_colour_manual(values = my_colours)+
  scale_fill_manual(values=my_colours)+
  theme_classic()+
  theme(legend.position = "none")+
  labs(x="",
       y="Distance Travelled (cm)")

y <- x + geom_jitter(data=data,aes(x=group,y=CHAL),size=5,alpha=0.5,width=0.15)

mean(data$CHAL)

Null <- ggplot(data=data, aes(x=ID,y=CHAL, colour=group))+
  geom_point(size=3,alpha=0.5)+
  scale_colour_manual(values = my_colours)+
  scale_fill_manual(values=my_colours)+
  theme_classic()+
  theme(plot.title = element_text(hjust=0.5))+
  theme(plot.subtitle = element_text(hjust=0.5))+
  theme(legend.position="none")+
  geom_hline(yintercept=mean(data$CHAL))+
  geom_segment(aes(x=ID,y=CHAL,xend=ID,yend=mean(data$CHAL)))+
  labs(x="ID",
       y="Distance travelled (cm)",
       title = "Model #1 (null model)",
       subtitle = "Horizontal Black line represents the mean \n of all the data points (a.k.a. the **Grand mean**)")+
  ylim(0,40000)

OIL_OIL <- data[data$group == "OIL / OIL", ]
EB_OIL <- data[data$group == "EB / OIL", ]
OIL_EB <- data[data$group == "OIL / EB", ]
EB_EB <- data[data$group == "EB / EB", ]

O.O_mean <- mean(OIL_OIL$CHAL)
E.O_mean <- mean(EB_OIL$CHAL)
O.E_mean <- mean(OIL_EB$CHAL)
E.E_mean <- mean(EB_EB$CHAL)

Exp <- ggplot(data=OIL_OIL, aes(x=ID,y=CHAL))+
  geom_point(colour="blue", size=3,alpha=0.5)+
  geom_point(data=EB_OIL, aes(x=ID,y=CHAL),colour="red",size=3,alpha=0.5)+
  geom_point(data=OIL_EB, aes(x=ID,y=CHAL),colour="green",size=3,alpha=0.5)+
  geom_point(data=EB_EB, aes(x=ID,y=CHAL),colour="purple",size=3,alpha=0.5)+
  geom_hline(yintercept=O.O_mean,colour="blue")+
  geom_hline(yintercept=E.O_mean,colour="red")+
  geom_hline(yintercept=O.E_mean,colour="green")+
  geom_hline(yintercept=E.E_mean,colour="purple")+
  geom_segment(data=OIL_OIL, aes(x=ID,y=CHAL,xend=ID,yend=O.O_mean),colour="blue")+
  geom_segment(data=EB_OIL, aes(x=ID,y=CHAL,xend=ID,yend=E.O_mean),colour="red")+
  geom_segment(data=OIL_EB, aes(x=ID,y=CHAL,xend=ID,yend=O.E_mean),colour="green")+
  geom_segment(data=EB_EB, aes(x=ID,y=CHAL,xend=ID,yend=E.E_mean),colour="purple")+
  theme_classic()+
  theme(plot.title = element_text(hjust=0.5))+
  theme(plot.subtitle = element_text(hjust=0.5))+
  labs(x="ID",
       y="Distance travelled (cm)",
       title = "Model #2 (experimental model)",
       subtitle = "Horizontal coloured lines represent the means \n of each of the 4 groups")+
  ylim(0,40000)

ggsave("Null.png", Null, height=4,width=4,dpi=300)
ggsave("Exp.png",Exp,height=4,width=4,dpi=300)
```

# One-Way ANOVA (Independent Samples)

```{r,echo=FALSE}
ggarrange(Null,Exp)
```

- Here, we are comparing whether the 4 individual **group means** is a better fit for the data than the **Grand Mean** 
- Vertical coloured lines represent the **residuals** 
- To begin calculating ANOVA, each **residual** is squared (^2)
- The **sum of squares** = the sum of all the squared residuals (All added together)

Mean Square is calculated separately for each model:

$$ Mean\ Square = \dfrac {Sum\ of\ Squares}{df} $$

The ratio between the two models is computed as the final *F* value:

$$ F =  \dfrac{Mean\ Square\ Between}{Mean\ Square\ Within} $$
### Prepare data 

- Prepare a column named "group" that codes the four experimental conditions.

```{r}
# create a single variable - "group" - to represent hormone treatments
data <- data %>%
  mutate(group = case_when(
    data$PREhorm == "OIL" & data$CHALhorm == "OIL" ~ "OIL / OIL",
    data$PREhorm == "EB" &  data$CHALhorm == "OIL" ~ "EB / OIL",
    data$PREhorm == "OIL" & data$CHALhorm == "EB" ~ "OIL / EB",
    data$PREhorm == "EB" &  data$CHALhorm == "EB" ~ "EB / EB"
  ))
# Reorder the "group" variable 
data$group <- as.character(data$group)
data$group <- factor(data$group, levels=c("OIL / OIL", "EB / OIL", "OIL / EB", "EB / EB"))
```

### Compute one-way ANOVA

```{r}
oneway.test(CHAL ~ group, data=data, var.equal = TRUE)
```

### Compute Tukey Follow-up comparisons of pairwise between-group comparisons. 

```{r}
a <- aov(CHAL ~ group, data=data)
TukeyHSD(a, conf.level = .95)
```

- The above example is not "the best" application of the OneWay ANOVA, or the best analysis for the data - The above approach would be appropriate if the 4 conditions were unrelated. 

- Here, the experiemnt is really a 2x2 design (PREhorm*CHALhorm), so it would be more appropriate to analyze using a Two-way ANOVA instead.

# Two-Way ANOVA on EB Rats Data  (Indpendent Samples)

```{r}
a <- aov(CHAL ~ PREhorm * CHALhorm, data=data)
summary(a)
```

## Written Result: 

> Rats treated with EB during the induction phase of sensitization travelled longer distances on the challenge day than rats treated with OIL vehicle during induction (*F*(1,43) = 22.40, *p* < 0.001).

> The effect of EB treatment on the challenge day and the interaction between induction and challenge hormone were non-significant (*p* = 0.39, *p* = 0.86, respectively).

# One-Way Repeated Measures ANOVA

Could be used to answer the question *do the rats exhibit overall increases in distance travelled across the multi-injection paradigm?* 

- The question could also be framed as *deos the nicotine paradigm produces locomotor sensitization?*

The same data are plotted to showcase individual changes in distance travelled across the four days of testing. 

```{r,echo=FALSE, fig.align='center'}
data %>%
  select(c("ID","Hab","IND_1","IND_2","CHAL")) %>%
  melt(id.vars = "ID") %>%
  ggplot(aes(x=variable,y=value,group=ID))+
  geom_line()+
  theme_classic()+
  theme(plot.title = element_text(hjust=0.5))+
  labs(x="Day of Testing",
       y="Distance Travelled (cm)",
       title = "Individual changes in Distance Across Days of Testing")
```

### Prepare Data
```{r} 
library(reshape2)
long_data <- data %>%
  select(c("ID","Hab","IND_1","IND_2","CHAL")) %>%
  melt(id.vars = "ID")
head(long_data)
```

### Compute ANOVA
```{r}
library(rstatix)
a <- anova_test(data=long_data, dv=value, wid=ID, within = variable)
get_anova_table(a)
```

# Compute Follow-up comparisons
```{r}
long_data %>%
  pairwise_t_test(
  value ~ variable, paired = TRUE,
  p.adjust.method = "bonferroni"
  )
```

## Written Result:

> Rats treated with EB during the induction phase travelled longer distances on the challenge day than did rats treated with OIL vehicle during induction (*F*(1,45) = 23.19, *p* < 0.001)

> Rats treated with EB during the induction phase exhibited an increase in distance travelled between the second induction day and the challange day (*p* < 0.001) whereas rats treated with OIL vehicle during induction did not (*p* = 0.24)

# Two-Way Repeated Measures ANOVA

### Prep Data 
- Output: A "long-form" version of the dataframe.
```{r} 
library(reshape2) # Required for the melt function
long_data_2 <- data %>% # Take the long dataframe AND TEHN 
  select(c("ID","PREhorm","Hab","IND_1","IND_2","CHAL")) %>% 
  melt(id.vars = c("ID","PREhorm")) # Switch to ggplot format
head(long_data_2) # Check out the result
```

### Compute ANOVA

```{r, out.width="50%"}
a <- anova_test(data=long_data_2, dv=value, wid=ID, within = variable, between = PREhorm)
# Make a pretty table: 
knitr::kable(get_anova_table(a))
```

### Compute Follow-up comparisons version 1 of 2:

- This style of follow up allows you to assess individual changes between days within individuals of different groups. 

```{r}
a <- long_data_2 %>%
  group_by(PREhorm)%>%
  pairwise_t_test(
  value ~ variable, paired = TRUE,
  p.adjust.method = "bonferroni"
  )
# Make a pretty table: 
library(kableExtra)
knitr::kable(a)
```

### Compute Follow-up comparisons version 2 of 2:

- This style of follow up allows you to assess between-group differences on each day of testing.

```{r}

a <- long_data_2 %>%
  group_by(variable)%>%
  pairwise_t_test(
  value ~ PREhorm, paired = FALSE,
  p.adjust.method = "bonferroni"
  )
# Make a pretty table: 
knitr::kable(a)
```
















