---
title: "Subsetting"
author: "JLB"
date: "2023-08-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Subsetting Data in R

- Subsequent data processing setps (statistical analysis, graph generation) works best if you select the data that you are interested in. Splitting up large data sets into multiple subsetted dataframes has many functions in R programming. 

# Get data
```{r}
library(tidyverse) 
data <- read_csv("EB_Rats_Nicotine_Sensitization.csv")
head(data)
```

# Select some of the columns

### Method #1: by position 

- In R language, square brackets [ ] mean "where". Inside of the square brackets, there can be two sets of numbers separated by a comma. Values that come before the comma refer to the ROWS and values that come after the comma refer to the COLUMNS. Blank space means "take all". 

- Square brackets can be used to select a "subset" of a larger dataframe. 

```{r}
# Select the first 6 columns 
data[ ,1:6] 

# You can also assign the subsetted data frame onto a new variable 
# It is generally better to create new variables than to overwrite. 
data_2 <- data[ ,1:6] 
head(data_2) 

# Select the first 6 rows 
data[1:6, ]

# Select the first 6 rows of the first 6 columns 
data[1:6,1:6]
```

### Method #2: by content

- Especially when working with larger datasets, you may want to subset based on cell values rather than position within the dataframe. Selecting based on content is a more robust approach because it does not require manual accuracy checks. 

```{r}
# Select all columns for the rows that have the value "OIL" in the column PREhorm:
data[data$PREhorm == 1, ]
```

```{r}
# Can also include multiple criteria: 
data[data$PREhorm == 1 & data$CHALhorm == 1, ]
```

### Method #3: dplyr "select"

- You can also use the "select" function inside of dplyr code blocks. This is generally the most elegant solution to choosing a subset of your dataframe to work with.

```{r}
data %>%
  select(c("ID","PREhorm","Hab"))
```





